
---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/README-",
  out.width = "100%"
)
```

# An example SEIR compartmental model implemented in stan

> [There are many like it, but this one is mine](https://en.wikipedia.org/wiki/Rifleman%27s_Creed)

An example SEIR model implemented in stan for the CMMID computational and inference theme meeting. 

## What

> Stan is a state-of-the-art platform for statistical modeling and high-performance statistical computation. Thousands of users rely on Stan for statistical modeling, data analysis, and prediction in the social, biological, and physical sciences, engineering, and business.

>Users specify log density functions in Stan’s probabilistic programming language and get:

> - full Bayesian statistical inference with MCMC sampling (NUTS, HMC)
> - approximate Bayesian inference with variational inference (ADVI)
> - penalized maximum likelihood estimation with optimization (L-BFGS)

> Stan’s math library provides differentiable probability functions & linear algebra (C++ autodiff). Additional R packages provide expression-based linear modeling, posterior visualization, and leave-one-out cross-validation.

See [here](https://mc-stan.org) for more. 

## Why

- Community, documention, and ongoing development.
- A good enough tool for a broad range of problems. Useful if wanting to do anything other than compartmental models.
- Clean DSL which can be extended with functions written in C++.
- Largely automated and optimised MCMC reduces the cognitive load when modelling.
- Number of stan users in the CMMID who are likely willing to help quid pro quo.

## Why not

- MCMC often not a great choice for complex compartmental model systems. For these models tools that support PMCMC or SMC^2 are likely more optimal (Libbi/Birch, ODIN/Dust, etc). 
- Not likely to be widely used by many Supervisors and so may be difficult to get code support. 
- Hard to use programmatically across models (you may find yourself writing a DSL generator).
- Interaction with fit model objects is not currently ideal but this is an area of work in the community (and the situation is much better than for other tools).

## Resouces (a small selection): 

- [Bayesian workflow for disease transmission modelling in stan](https://mc-stan.org/users/documentation/case-studies/boarding_school_case_study.html)
- [Fitting Bayesian models using Stan and R](https://www.weirdfishes.blog/blog/fitting-bayesian-models-with-stan-and-r/)
- [Fitting a basic SIR model in stan](https://www.generable.com/blog/2020/04/fitting-a-basic-sir-model-in-stan/)
- [Contemporary statistical inference for infectious disease models using Stan](https://arxiv.org/abs/1903.00423v3)
- [covidseir](https://github.com/seananderson/covidseir)

## Example

Code below is heavily based on [Bayesian workflow for disease transmission modelling in stan](https://mc-stan.org/users/documentation/case-studies/boarding_school_case_study.html).


### Set up

* Load some packages

```{r, message = FALSE}
library(rstan)
library(tidyverse)
```

* Load the model

```{r}
model <- stan_model("model.stan")
model
```

* Load the data

```{r}
# Cases
dt <- readRDS("data/SIRsample.rds")
head(dt)
```

* Look at the simulated data (it is a mess - thanks Carl)

```{r}
plot(dt$onsets)
```
* Define as parameters for the stan model

```{r}
data <- list(
  cases = dt$onsets,
  n_days = length(dt$onsets),
  t0 = 0,
  tswitch = 10,
  N = 1e5
)
data$ts <- seq(1, data$n_days, by = 1)
```

### Fit the model


* Fit the model using NUTS

```{r, eval = FALSE}
fit_nuts <- sampling(model,
                     data = data,
                     chains = 4,
                     seed = 0)
```